---
title: "hw2 jw4007"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

```{r load_libraries}
library(tidyverse)
library(readxl)
```

### Prblem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "./NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%
  janitor::clean_names() %>%
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

trans_ent
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```
The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 
```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% pull(entry) %>% 
  mean
```
Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>%  
  select(station_name, line) %>% 
  distinct
  
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct

trans_ent
  
```

### Problem 2

#### 1) Read and clean the Mr. Trash Wheel sheet:
```{r}
mr_trash = read_excel("./Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N549") %>%
  janitor::clean_names() %>%
  filter(dumpster != "NA") %>%
  mutate(year = as.numeric(year)) %>%
  mutate(sports_balls = as.integer(sports_balls), who = "mr")

mr_trash_2020 = filter(mr_trash, year == 2020) 
```

The "Mr. Trash Wheel" dataset has `r nrow(mr_trash)` observations with the following variables: `r colnames(mr_trash)`.The total number of sports balls collected by Mr. Trash Wheel in 2020 is `r sum(pull(mr_trash_2020, sports_balls))`.

#### 2) Read and clean the Professor Trash Wheel sheet:
```{r}
prof_trash = read_excel("./Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M96") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(who = "prof")

sum(prof_trash$weight_tons)
```
The "Professor Trash Wheel" dataset has `r nrow(prof_trash)` observations with the following variables: `r colnames(prof_trash)`.The total weight of trash collected by Professor Trash Wheel is `r sum(pull(prof_trash, weight_tons))`.

#### 3) Combine the two to produce a single tidy dataset
```{r}
mr_prof = full_join(mr_trash, prof_trash)
```
The final dataset has `r nrow(mr_prof)` observations with the following variables: `r colnames(mr_prof)`.


### Problem 3
#### 1) clean the pols-month:
```{r}
pols_month = 
  read_csv(
    "./q3data/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(month = as.numeric(month), month = month.abb[month], month = str_to_lower(month)) %>% 
  mutate(president = prez_gop + prez_dem) %>%
  mutate(year = as.numeric(year)) %>%
  select(-prez_gop, -prez_dem, -day) %>%
  arrange(year, month)

max(pull(pols_month, year))
min(pull(pols_month, year))
```
The pols_month dataset has `r nrow(pols_month)` rows and `r ncol(pols_month)` columns with the following variables: `r colnames(pols_month)`. The years ranges from `r min(pull(pols_month, year))` to `r max(pull(pols_month, year))`

#### 2) clean the snp:
```{r}
snp = read_csv(
    "./q3data/snp.csv") %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(month = as.numeric(month), month = month.abb[month], month = str_to_lower(month)) %>%
  mutate(year = as.numeric(year), year = 2000 + year) %>%
  select(-day) %>%
  arrange(year, month)
  
```
The snp dataset has `r nrow(snp)` rows and `r ncol(snp)` columns with the following variables: `r colnames(snp)`. The years ranges from `r min(pull(snp, year))` to `r max(pull(snp, year))`

#### 3) tidy the unemployment:
```{r}
unemploy = read_csv(
    "./q3data/unemployment.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(jan:dec, names_to = "month", values_to = "percent_unemploy")
  
```
The unemploy dataset has `r nrow(unemploy)` rows and `r ncol(unemploy)` columns with the following variables: `r colnames(unemploy)`. The years ranges from `r min(pull(unemploy, year))` to `r max(pull(unemploy, year))`

#### 4) joining the datatsets:
```{r}
pols_snp = full_join(pols_month, snp, by = c("year" = "year", "month" = "month"))

unemploy_pols_snp = full_join(pols_snp, unemploy, by = c("year" = "year", "month" = "month"))
```
The final dataset combining unemploy, snp, and pols_month has `r nrow(unemploy_pols_snp)` rows and `r ncol(unemploy_pols_snp)` columns with the following variables: `r colnames(unemploy_pols_snp)` (dimension:`r dim(pull(unemploy_pols_snp))`). The years ranges from `r min(pull(unemploy_pols_snp, year))` to `r max(pull(unemploy_pols_snp, year))`


